Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
own extension, overall algorithm

%likely shouldn't be an entire section
\section{EAGGA}
EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
The three interpretability proxies are
\begin{itemize}
  \item NF: the relative number of features used in the model
  \item NI: the relative number of pairwise interaction effects in the model
  \item NNM: the relative number of non-monotone feature effects in the model
\end{itemize}
The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
translate to a higher chance of approval.
Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
\\
\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the large objective space $\check{\boldsymbol\Lambda}$ to be optimized over, containing
tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
with
\begin{itemize}
  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
\end{itemize}
Consequently, the optimization problem over this space would be
\begin{equation}
  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
  \big)
  \label{eq-eagga-extended-space}
\end{equation}
The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
Using this group structure, they define a much simpler augmented serach space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
\begin{itemize}
  \item $G_1=C\setminus C_s$ being the set of excluded features
  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
  \begin{itemize}
    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
    \item $M_{E_k}$ being the monotonicity constraint of group $k$
  \end{itemize}
\end{itemize}
To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
\citep[pp. 541-543]{EAGGA}

\section{Extension to neural networks}
- original EAGGA applied to XGBoost model proved to vastly outperform union of competitor models w.r.t. dominated hypervolume and comparable or better than ParEGO on
  extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}
- why extend it to NNs?
  * NNs notoriously uninterpretable due to complex transformation of the feature space
  * due to (lin alg) non-linear activation functions, interactions can be modelled
  * no monotonicity guarantees
- our approach
  (general algorithm)
  * EAGGA algorithm implemented largely as described in the paper
  * HP init
    - NN total layers $\in\{3, ..., 10\}$, i.e. hidden layers $\in\{1, ..., 8\}$, init from trunc geom with prob=0.5
    - NN nodes per hidden layer (for each group) $\in\{3, ..., 20\}$, init from trunc geom with prob=0.5
    - NN dropout \% $\in[0, 1]$, init from trunc gamma with shape=2, scale=0.15
  * group structure init
    - feature detector
      * as described in paper
      * but instead of fitting 10 trees + taking rel. \# of features used as prob for trunc geom, we simply use 0.5 (sklearn dectree examination not straightforward)
      * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we simply use all features with non-0 filter values
    - interaction detector
      * as described in paper
      * same as for feat detect: use prob=0.5 to sample \# of interactions used instead of 10 dec trees
      * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
    - monotonicity detector
      * as described in paper
      * use default HPs (max depth 30, minsplit = 20) of mlr3 classification tree implementation, as this is the library the paper uses
    -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
  (implementation details)

@@

      * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN

@@

  * our implementation of feature + interaction detectors likely tend to include

@@

    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)

@@

@@

\newpage

>>>>

%likely shouldn't be an entire section
\section{EAGGA}
EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
The three interpretability proxies are
\begin{itemize}
  \item NF: the relative number of features used in the model
  \item NI: the relative number of pairwise interaction effects in the model
  \item NNM: the relative number of non-monotone feature effects in the model
\end{itemize}
The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
translate to a higher chance of approval.
Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
\\
\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the complex ``extended search space'' $\check{\boldsymbol\Lambda}$ to be optimized over, containing
tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
with
\begin{itemize}
  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
\end{itemize}
Consequently, the optimization problem over this space would be
\begin{equation}
  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
  \big)
  \label{eq-eagga-extended-space}
\end{equation}
The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
Using this group structure, they define a much simpler ``augmented serach space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
\begin{itemize}
  \item $G_1=C\setminus C_s$ being the set of excluded features
  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
  \begin{itemize}
    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
    \item $M_{E_k}$ being the monotonicity constraint of group $k$
  \end{itemize}
\end{itemize}
To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
\citep[pp. 541-543]{EAGGA}

\section{Extension to Neural Networks}
EAGGA as implemented by \citet{EAGGA} applies the optimizer to XGBoost, which proved to vastly outperform a union of competitor models with respect to
dominated hypervolume and achieved comparable or better performance than ParEGO optimizing over the
extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}.
Our work extends EAGGA to neural networks. We chose this model class for several reasons:
\begin{enumerate}[label*=\arabic*.]
  \item NNs are notoriously uninterpretable due to complex transformations of the feature spaces
  \item Non-linear activation functions implicitly model interactions without direct control of which features interact with each other
  \item There are no monotonicty guarantees
\end{enumerate}

\subsection{Overall Algorithm}
We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
\end{itemize}
For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

\subsection{Architectural Details}

@@

      * comprises "sub-NNs", one for each equivalence relation -> basically non-fully connected NN

@@

  * our implementation of feature + interaction detectors likely tend to include  % belongs in "general algorithm"

@@

    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constrained)

@@

    % TODO: put explanation in line 400, don't introduce new topic here

@@

%\newpage  % we are not to add page breaks
